*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-21 Code/Manuscript_O2.Rmd
Added more citations, fixed order of results figures

XXXXMIREILLE Check strain descriptions for accuracyXXXX
Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral ranges [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, at depths from 120 m and 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021] (Lavin et al., 2010) ADDITIONAL REFS XXXX.   (Johnson et al., 1999). @goerickeNovelNicheProchlorococcus2000
[@garcia-robledoCrypticOxygenCycling2017a]. *P. marinus* LL ecotypes may indeed dominate the phytoplankton wihthin OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021]
[@partenskyProchlorococcusAdvantagesLimits2010].

@@

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are  typically dominant picophytoplankton in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI, including cultured strain MED4,is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface (Partensky et al., 1999).  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant group in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population (Partensky et al., 1999),  and are most numerous around 50 m depth (Partensky et al., 1999). Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

*P. marinus* clades are found  in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes
[@westNichePartitioningProchlorococcusPopulationsStratified1999;
@zinserInfluenceLightTemperature2007;
@delmontLinkingPangenomesMetagenomes2018] 
[@moorePhotophysiologyMarineCyanobacterium1999]. 
@lavinNovelLineagesProchlorococcus2010 and
@mattisonPhotosyntheticStrategiesWild2020 show that LL ecotypes
can occupy regions in the OMZ at depths above 40 m, reaching ambient light levels above what LL clades were thought to tolerate.

@@

Under full atmospheric [O~2~], LL clades of *Prochlorococcus* are restricted to growth under low light, in part because they suffer photoinhibition of Photosystem II (PSII) through several paths, including direct absorbance of UV or blue light, in parallel with generation of Reactive Oxygen Species (ROS) if the electron flow is slowed, [@aroPhotoinhibitionPhotosystemII1993], producing  damaging singlet oxygen (^1^O~2~). [@aroPhotoinhibitionPhotosystemII1993; @soitamoPhotoinhibitionMarinePicocyanobacteria2017] [@murphyPhotoinactivationPhotosystemII2017][@hakalaEvidenceRoleOxygenevolving2005].
Repair of inhibited PSII process relies on the removal of damaged PsbA (Mann et al., 2000; Adam et al., 2005), followed by reassembly with newly synthesized PsbA [@nixonRecentAdvancesUnderstanding2010]. 
Thus, degradation of PsbA is a rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993), mediated largely by a heterohexamer of (FtsH12)3, a membrane-bound (Sacharz et al., 2015)(Zak et al., 2001) metalloprotease  (Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014).  (Adam et al., 2005; Boehm et al., 2012) (Nixon et al. (2010)) (Komenda et al., 2007). (Pisareva et al., 2007; Sacharz et al., 2015),  (Mann et al., 2000), (AMANDA REF from ALGATECH).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4, homologous to the characterized FtsH isoforms of the model cyanobacterium *Synechocystis* FtsH, with presumably parallel functions (Table 1).  Upon a shift to higher light HLIII MED4 upregulated expression of FtsH1 and FtsH2 (Bonisteel et al. (2018)), homologs to the *Synechocystis* slr0228 and slr1604, implicated in PSII repair. In contrast representative LLIV strain MIT9313 showed less overall expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem.  Furthermore MIT9313 expressed primarily FtsH3, homologous to Sll1463 involved in PSI biogenesis, and FtsH expression did not increase in response to light stress. Through adaptation to steady low light, LLIV *Prochlorococcus* instead allocates resources processes other than dynamic regulation of PSII repair.

@@

## Growth data management and analysis {.unnumbered}
Data files (.csv) saved from the Multicultivator software were imported into R-Studio [@R-Studio] for data management [@R-tidyverse], growth rate calculations CITATION, comparisons of model fits CITATION, and generation of
figures using `ggplot2` [@R-ggplot2]. The chlorophyll proxy
optical density (OD680 - OD720;  ΔOD) was used to determine the chlorophyll-specific exponential growth
rate (µ, d^-1^) for each treatment combination.  We first we used a rolling mean from the RStudio `zoo` package [@R-zoo] to calculate the average ΔOD data over a 1-hour window, to lower influence of outlier points and remove data points collected during post stationary phase, if applicable. We used the Levenberg-Marquardt algorithm
[@bellaviaLevenbergMarquardtMethod2018] modification of the non-linear
least squares,  using the R package `minpack.lm`
[@R-minpack.lm], to fit a logistic
equation \@ref(eq:GrowthRate) where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.  Annotated code for data import, transformations and analyses is available at XXXXX 
CITAtIONS of github locations of code etc. XXXXX

@@

PurParPlot_cap <- glue("**Absorbed peak Photosynthetically Usable Radiation (PUR) (µmol photons m^-2^ s^-1^) vs. peak Photosynthetically Active Radiation (PAR) (µmol photons m^-2^ s^-1^).**",
                       "The correlation between PAR, plotted on the x-axis and absorbed PUR, plotted on the y-axis, are colored for each growth spectral waveband; blue for 450 nm, red for 660 nm and black for white LED full spectrum light.",
                       "The grey dashed line represents a hypothetical one to one correlation.",
                       "**A.** is *Prochlorococcus marinus*  MED4.",
                       "**B.** is *Prochlorococcus marinus*  SS120.",
                       "**C.** is *Prochlorococcus marinus*  MIT9313.",

@@

# Results and discussion {.unnumbered}

@@

## Generalized Additive Models (GAM) of *Prochlorococcus marinus* growth responses to photoperiod, PAR, spectal bands and O~2~   {.unnumbered}

### SS120 Clade Clade LLIII  {.unnumbered}

@@

@@

The representative of HLI clade, *P. marinus* MED4, was completely
excluded from growth under the 4 h photoperiod, andshowed
increasing growth rates with increasing photoperiod. Thus, MED4 can
potentially exploit longer summer photoperiods in temperate zones, but
will likely remain excluded from growth in winter through its requirement for
more than 4 h of light per day, even if water temperature warms into the Clade HLI tolerance range.

@@

#### MED4 Clade HLI {.unnumbered}

#### SS120 Clade LLIII {.unnumbered}

#### MIT9313 Clade LLIV {.unnumbered}

## Discussion PUR and growth trends {.unnumbered}

@@

## Discussion from Mireille' Thesis General Discussion section {.unnumbered}

@@

In chapter 2, we analyzed  growth rates to determine the viability of *P. marinus* MED4, a clade HLI ecotype found at depths near the ocean surface,
*P. marinus* SS120, a clade LLIII ecotype found deep in the water column, and *P. marinus* MIT9313, a clade LLIV ecotype also found in deep oceans, under present day and hypothetical future niche conditions. We designed our study to account for light attenuation through ocean waters, recognizing that low light levels of the blue spectral waveband ultimately prevail in deep ocean waters [@morelAvailableUsableStored1978; @holtropVibrationalModesWater2021] and that the solar incidence angle also attenuates the photoperiod or hours of light exposure beneath the ocean surface. We also tested the growth response of *P. marinus* to varying [O~2~] under the matrix of light parameters to determine if *P. marinus* can cope with  diverse  ecosystem scenarios.

In chapter 3, we determined the Photosynthetic Usable Radiation (PUR), the portion of light radiation absorbed differentially by the pigments of the photosynthetic organism [@morelAvailableUsableStored1978] and expressed growth response in terms of cumulative diel PUR.  
We found that approximately 1.0 x 10^6^ µmol photons m^-2^ d^-1^ of PUR is the maximum dosage that the representative LL clades, SS120 and MIT9313 can tolerate under 25 µM O~2~, in contrast to the representative HL clade, MED4 which exhibited growth saturation at the same cumulative diel PUR of 1.0 x 10^6^ µmol photons m^-2^ d^-1^.  Growth under lower O~2~ allowed MIT9313 to substantially increase its exploitation of higher diel PUR.

XXX ProchlorococcusEnzymeKms figure was here XXX

@@

XXX DNARepairFig was here XXX

%%%% OceanFtsHRibosomeProteins figure from OceanPortal repository  %%%

>>>>

XXXXMIREILLE Check strain descriptions for accuracyXXXX
Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral ranges [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, at depths from 120 m and 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @@lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017a] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010].

@@

HL clades are the most recently evolved lineages, with reduced genome sizes in comparison to LL ecotypes. High-Light clades are  typically dominant picophytoplankton in near-surface, oligotrophic waters, characterized by high light levels. Clade HLI, including cultured strain MED4,is adapted to higher iron, and lower temperatures, and originated from the Mediterranean Sea, just 5 m below the surface [@partenskyProchlorococcusMarinePhotosynthetic1999].  Clade HLII adapted to higher iron, and higher temperatures, is the most abundant group in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population [@partenskyProchlorococcusMarinePhotosynthetic1999],  and are most numerous around 50 m depth [@partenskyProchlorococcusMarinePhotosynthetic1999]. Clade HLIII/IV is adapted to lower iron [@kentParallelPhylogeographyProchlorococcus2019; @johnsonNichePartitioningProchlorococcus2006; @zinserInfluenceLightTemperature2007].

*Prochlorococcus marinus* clades are found  in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes
[@westNichePartitioningProchlorococcusPopulationsStratified1999;
@zinserInfluenceLightTemperature2007;
@delmontLinkingPangenomesMetagenomes2018] 
[@moorePhotophysiologyMarineCyanobacterium1999]. 
@lavinNovelLineagesProchlorococcus2010 and
@mattisonPhotosyntheticStrategiesWild2020 show that LL ecotypes
can occupy regions in the OMZ at depths above 40 m, reaching ambient light levels above what LL clades were thought to tolerate.

@@

@@

## Growth data management and analysis {.unnumbered}
Data files (.csv) saved from the Multicultivator software were imported into R-Studio [@R-Studio] for data management [@R-tidyverse], growth rate calculations CITATION, comparisons of model fits (get Platt CITATION), and generation of
figures using `ggplot2` [@R-ggplot2]. The chlorophyll proxy
optical density (OD680 - OD720;  ΔOD) was used to determine the chlorophyll-specific exponential growth
rate (µ, d^-1^) for each treatment combination.  We first we used a rolling mean from the RStudio `zoo` package [@R-zoo] to calculate the average ΔOD data over a 1-hour window, to lower influence of outlier points and remove data points collected during post stationary phase, if applicable. We used the Levenberg-Marquardt algorithm
[@bellaviaLevenbergMarquardtMethod2018] modification of the non-linear
least squares,  using the R package `minpack.lm`
[@R-minpack.lm], to fit a logistic
equation \@ref(eq:GrowthRate) where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.  Annotated code for data import, transformations and analyses is available at XXXXX 
CITAtIONS of github locations of code etc. XXXXX

@@

PurParPlot_cap <- glue("**Absorbed peak Photosynthetically Usable Radiation (PUR) (µmol photons m^-2^ s^-1^) vs. peak Photosynthetically Active Radiation (PAR) (µmol photons m^-2^ s^-1^).**",
                       "The correlation between PAR, plotted on the x-axis and absorbed PUR, plotted on the y-axis, are colored for each growth spectral waveband; 450 nm (blue circles), 660 nm (red circles) and white LED (black circles).",
                       "The grey dashed line represents a hypothetical one to one correlation.",
                       "**A.** is *Prochlorococcus marinus*  MED4.",
                       "**B.** is *Prochlorococcus marinus*  SS120.",
                       "**C.** is *Prochlorococcus marinus*  MIT9313.",

@@

# Results {.unnumbered}

@@

@@

Under full atmospheric [O~2~], LL clades of *Prochlorococcus* are restricted to growth under low light, in part because they suffer photoinhibition of Photosystem II (PSII) through several paths, including direct absorbance of UV or blue light, in parallel with generation of Reactive Oxygen Species (ROS) if the electron flow is slowed, [@aroPhotoinhibitionPhotosystemII1993], producing  damaging singlet oxygen (^1^O~2~) [@aroPhotoinhibitionPhotosystemII1993; @soitamoPhotoinhibitionMarinePicocyanobacteria2017; @murphyPhotoinactivationPhotosystemII2017; @hakalaEvidenceRoleOxygenevolving2005].
Repair of inhibited PSII process relies on the removal of damaged PsbA (Mann et al., 2000; Adam et al., 2005), followed by reassembly with newly synthesized PsbA [@nixonRecentAdvancesUnderstanding2010]. 
Thus, degradation of PsbA is a rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993), mediated largely by a heterohexamer of (FtsH12)3, a membrane-bound (Sacharz et al., 2015)(Zak et al., 2001) metalloprotease  (Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014).  (Adam et al., 2005; Boehm et al., 2012) (Nixon et al. (2010)) (Komenda et al., 2007). (Pisareva et al., 2007; Sacharz et al., 2015),  (Mann et al., 2000), (AMANDA REF from ALGATECH).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4, homologous to the characterized FtsH isoforms of the model cyanobacterium *Synechocystis* FtsH, with presumably parallel functions (table \@ref(tab:FtsHTable)). Upon a shift to higher light HLIII MED4 upregulated expression of FtsH1 and FtsH2 [@bonisteelStrainSpecificDifferences2018], homologs to the *Synechocystis* slr0228 and slr1604, implicated in PSII repair. In contrast representative LLIV strain MIT9313 showed less overall expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem.  Furthermore MIT9313 expressed primarily FtsH3, homologous to Sll1463 involved in PSI biogenesis, and FtsH expression did not increase in response to light stress. Through adaptation to steady low light, LLIV *Prochlorococcus* instead allocates resources processes other than dynamic regulation of PSII repair.

# Discussion {.unnumbered}

@@

The representative of HLI clade, *P. marinus* MED4, was completely
excluded from growth under the 4 h photoperiod, and showed
increasing growth rates with increasing photoperiod. Thus, MED4 can
potentially exploit longer summer photoperiods in temperate zones, but
will likely remain excluded from growth in winter through its requirement for
more than 4 h of light per day, even if water temperature warms into the Clade HLI tolerance range.

@@

### Discussion PUR and growth trends {.unnumbered}

@@

### Discussion from Mireille' Thesis General Discussion section {.unnumbered}

@@

@@
